---
title: "An R Notebook Exploring Green Cab Data"
output: html_notebook
---

Using data from: http://www.nyc.gov/html/tlc/html/about/trip_record_data.shtml, we explore all trips taken in Green Cabs in New York City during January 2016.

```{r get_data}
library(tidyverse)
library(htmlwidgets)
library(lubridate)
library(magrittr)
library(data.table)

taxi <- read_csv("data/green_tripdata_2016-01.csv")

setDT(taxi)
taxi$trip_time <- taxi$Lpep_dropoff_datetime - taxi$lpep_pickup_datetime

# minimal data cleaning
# remove very long (> 5 hrs) and very short trips (< 1 min)
taxi <- taxi[taxi$trip_time > 60,]
taxi <- taxi[taxi$trip_time < 18000,]
```


### When are people hailing Green Cabs? 

The following chart looks at pickup time by time of day and day of week

```{r time_of_day_of_week}
library(plotly)
p <- taxi[, .N, .(hour(lpep_pickup_datetime), weekdays(lpep_pickup_datetime))] %>% 
      ggplot(aes(x = hour, y = N)) + geom_bar(stat = "identity") + 
      facet_grid(factor(weekdays, 
                        levels = c("Sunday","Monday","Tuesday","Wednesday",
                                   "Thursday", "Friday","Saturday"),
                        labels = c("Sun","Mon","Tues","Wed","Thur","Fri","Sat")) ~ .) +
      ggtitle("Green Cab rides by day of week and time of day") + 
      xlab("Hour of day") + 
      theme_bw()

ggplotly(p)
```

## When do drivers get fares?

```{r dygraphs}
library(dygraphs)
library(xts)

taxi[, .N, as.IDate(lpep_pickup_datetime)] %>% as.xts() %>% 
  dygraph(main = "Number of Rides Taken in Green Cabs: Jan 2016") %>% 
  dyAxis("y", valueRange = c(0, 70000)) %>% 
  dyRangeSelector()

```

## When do drivers get tips?

```{r heatmap}
library(d3heatmap)

taxi[, round(mean(Tip_amount), digits = 2), .(hour(lpep_pickup_datetime), wday(lpep_pickup_datetime))] %>% 
  spread(key = wday, value = V1) %>% .[,c(2:8), with = F] %>% 
  d3heatmap(dendrogram = "none", 
            labCol = c("Sun","Mon","Tues","Wed","Thur","Fri","Sat"),
            colors = "Blues")

```

## Where do the big tips come from?

Occasionally someone gives a tip of $100 or more.  In January 2016 there were `r nrow(taxi[Tip_amount >= 100])` such rides in Green Cabs.  Where are these people getting picked up?  Mouseover for the trip distance and duration.

```{r leaflet}
library(leaflet)

# create a tidy data set
big_tip <- taxi[Tip_amount >= 100, c("Pickup_latitude","Pickup_longitude","lpep_pickup_datetime",
                                     "Trip_distance","trip_time")] %>% 
  setnames(old = c("Pickup_latitude","Pickup_longitude","lpep_pickup_datetime"),
           new = c("lat", "long","pickup"))

# Map it!
leaflet(big_tip) %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  setView(lng = -73.95, lat = 40.7, zoom = 10) %>%
  addMarkers(label = ~paste0("Distance: ",as.character(Trip_distance),
                             ", Time: ", as.character(round(trip_time/60, digits = 0)), 
                             "min, Hour of pickup: ", as.character(hour(pickup))))

```

## Which areas aren't tipping?

Let's see if there are places where you might (or might not) want to pick someone up if you wanted to maximize your tips.

```{r tipmap}
# make a 1% sample data set
set.seed(9486)
tipmap <- taxi[sample(x = c(T,F), size = nrow(taxi), replace = T, prob = c(0.01,0.99)),
               c("Pickup_latitude","Pickup_longitude","lpep_pickup_datetime",
                 "Trip_distance","trip_time","Tip_amount")] %>% 
            setnames(old = c("Pickup_latitude","Pickup_longitude","lpep_pickup_datetime"),
                     new = c("lat", "long","pickup"))

# make a color palette
quantile(tipmap$Tip_amount, c(0, 0.5, 0.6, 0.7, 0.8, 0.9, 0.95, 1))
tipmap$tip_cat <- cut(x = tipmap$Tip_amount, c(0,0.01,1,2,4,10,1500), include.lowest = T,
    labels = c("$0", "<$1", "$1.01-$2", "$2.01-$4", "$4.01-$10", ">$10"))
pal <- colorFactor(palette = c("#fc8d59","#fdcc8a","#b2e2e2","#66c2a4","#2ca25f","#006d2c"), 
                   domain = c("$0", "<$1", "$1.01-$2", "$2.01-$4", "$4.01-$10", ">$10"),
                   ordered = T)

leaflet(tipmap) %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  setView(lng = -73.95, lat = 40.7, zoom = 10) %>%
  addCircleMarkers(radius = 3, fillOpacity = 0.5, 
                   color = ~pal(tip_cat), stroke = FALSE,
                   label = ~paste0("$", Tip_amount))
```
